---
title: "&nbsp;"
output:
  html_document: default
  pdf_document: default
---

<center><h1>R Code for Climate Mathematics:</h1></center>
<center><h1>Theory and Applications</h1></center>
<p>&nbsp;</p>
<center><h3>A Cambridge University Press book By</h3></center>
<center><h3>SSP Shen and RCJ Somerville</h3></center>

<p>&nbsp;</p>
<p>&nbsp;</p>

<div align="center">Version 1.0 released in July 2019 and coded by Dr. Samuel Shen, Distinguished Professor</div>
<div align="center">San Diego State University, California, USA</div> 
<div align="center"><https://shen.sdsu.edu></div> 
<div align="center">Email: <sshen@sdsu.edu></div> 
<p>&nbsp;</p>
<div align="center">Version 2.0 compiled and updated by Momtaza Sayd</div> 
<div align="center">San Diego State University May 2021</div>
<div align="center">Version 3.0 compiled and updated by Joaquin Stawsky</div>
<div align="center">San Diego State University June 2022</div>

<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
## Chapter 2: Basics of R Programming

### R as a smart calculator
```{r}
1+4
2+pi/4-0.8
x <- 1
y <- 2
z <- 4
t <- 2*x^y-z
t
u = 2        # "=" sign and "<-" are almost equivalent
v = 3        # The text behind the "#" sign is comments
u+v
sin(u*v)    # u*v = 6 in the sine function is considered a radian by R 
```

<p>&nbsp;</p>
### Define a sequence in R
```{r}
#Enter temperature data in c() 
tmax <- c(77, 72, 75, 73, 66, 64, 59)
#Show the data
tmax
#Generate same sequence using different methods
seq(1,8)
seq(8)
seq(1,8, by = 1)
seq(1,8, length = 8)
seq(1,8, length.out = 8)
```


<p>&nbsp;</p>
### Define a function in R
```{r}
#Define a function
samfctn <- function(x) x*x
samfctn(4)
fctn2 <- function(x,y,z) x+y-z/2
fctn2(1,2,3)
```


<p>&nbsp;</p>
### Plot with R
```{r}
#Plot temperature data
plot(1:7, c(77, 72, 75, 73, 66, 64, 59))

#More plot examples
plot(sin, -pi, 2*pi)   #plot the curve of y = sin(x) from -pi to 2 pi

square <- function(x) x*x   #Define a function
plot(square, -3, 2)   # Plot the defined function

# Plot a 3D surface
x <- seq(-1, 1, length = 100)
y <- seq(-1, 1, length = 100)
z <- outer(x, y, function(x, y)(1-x^2-y^2))  
# 'outer(x,y, function)' renders z function on the x, y grid
persp(x,y,z, theta = 330) 
# yields a 3D surface with perspective angle 330 deg

#Contour plot
contour(x,y,z) #lined contours
filled.contour(x,y,z) #color map of contours with default colors
```


<p>&nbsp;</p>
### Symbolic calculations by R
```{r}
D(expression(x^2,'x'), 'x') 
# Take derivative of x^2 w.r.t. x 
2 * x #The answer is 2x

fx <- expression(x^2,'x')  #assign a function  
D(fx,'x') #differentiate the function w.r.t. x
2 * x  #The answer is 2x

fx <- expression(x^2*sin(x),'x') 
#Change the expression and use the same derivative command
D(fx,'x')

2 * x * sin(x) + x^2 * cos(x)

fxy <- expression(x^2+y^2, 'x','y') 
#One can define a function of 2 or more variables
fxy #Renders an expression of the function in terms of x and y
#expression(x^2 + y^2, "x", "y")
D(fxy,'x') #yields the partial derivative with respect to x: 2 * x
D(fxy,'y') #yields the partial derivative with respect to y: 2 * y

square <- function(x) x^2
#Integrate x^2 from 0 to 1 equals to 1/3 with details below
integrate(square,0,1)
#0.3333333 with absolute error < 3.7e-15

#Integrate cos(x) from 0 to pi/2 equals to 1 with details below
integrate(cos,0,pi/2) 
#1 with absolute error < 1.1e-14

```


<p>&nbsp;</p
### Vectors and matrices
```{r}
c(1,6,3,pi,-3) #c() gives a vector and is considered a 5X1 column vector
seq(2,6) #Generate a sequence from 2 to 6
seq(1,10,2) # Generate a sequence from 1 to 10 with increments of 2
x <- c(1,-1,1,-1)
x+1 #1 is added to each element of x
2*x #2 multiplies each element of x
x/2 # Each element of x is divided by 2
y <- seq(1,4)
x*y  # This multiplication * multiples each pair of elements 
x%*%y #This is the dot product of two vectors and yields a 1x1 matrix
t(x)  # Transforms x into a row 1X4 vector
t(x)%*%y #This is equivalent to dot product and forms 1X1 matrix
x%*%t(y) #This column times row yields a 4X4 matrix
my <- matrix(y,ncol = 2) 
#Convert a vector into a matrix of the same number of elements
#The matrix elements go by column, first column, second, etc
#Commands matrix(y,ncol = 2, nrow = 2)  or matrix(y,2)  
#or matrix(y,2,2) does the same job
my
dim(my)  #find dimensions of a matrix
as.vector(my) #Convert a matrix to a vector, again via columns
mx <- matrix(c(1,1,-1,-1), byrow = TRUE,nrow = 2)
mx*my #multiplication between each pair of elements
mx/my #division between each pair of elements
mx-2*my  
mx%*%my #This is the real matrix multiplication in matrix theory
det(my) #determinant
myinv <- solve(my) #yields the inverse of a matrix
myinv
myinv%*%my #verifies the inverse of a matrix
diag(my) #yields the diagonal vector of a matrix
myeig <- eigen(my) #yields eigenvalues and unit eigenvectors
myeig
myeig$values
myeig$vectors
mysvd <- svd(my) #SVD decomposition of a matrix M <- UDV'
#SVD can also be done for a rectangular matrix of mXn dimensions
mysvd$d
mysvd$u
mysvd$v

ysol <- solve(my,c(1,3)) 
#solve linear equations for matrix%*%x = b
ysol  #solve(matrix, b)
my%*%ysol #verifies the solution
```


<p>&nbsp;</p>
### Simple statistics by R
```{r}
x <- rnorm(10) #generate 10 normally distributed numbers
x
mean(x)

var(x)

sd(x)

median(x)

quantile(x)

range(x) #yields the min and max of x

max(x)

boxplot(x) #yields the box plot of x

w <- rnorm(1000)
hist(w) 
#yields the histogram of 1000 random numbers with a normal distribution

summary(rnorm(12)) #statistical summary of the data sequence

#Linear regression and linear trend line
#2007-2016 data of the global temperature anomalies
#Source: NOAAGlobalTemp data
t <- 2007:2016
T <- c(.36,.30, .39, .46, .33, .38, .42, .50, .66, .70)
lm(T ~ t) #Linear regression model of temp vs time

#Temperature change rate is 0.03673  deg C/yr or 0.37 deg C/decade
# expression(paste(...)) is one way of using the degree symbol
# ...pasting the ° symbol directly is another way
plot(t,T, type = "o",xlab = "Year",ylab = expression(paste("Temperature [",~degree,"C]")), 
     main = expression(paste("2007-2016 Global Temperature Anomalies\n and Their Linear Trend [0.37 °C/decade]")))
abline(lm(T ~ t), lwd = 2, col = "red") #Regression line
```


<p>&nbsp;</p>
### YouTube tutorial: Input data by reading a csv file into R
```{r}
#The R packages and the datasets used in this book are 
#listed below and can be downloaded and installed first 
#before proceeding to the R codes in the rest of the book. 
#The R packages: 
#animation, chron, e1071, fields, ggplot2, lattice, 
#latticeExtra, maps, mapdata, mapproj, matrixStats, ncdf, 
#NLRoot, RColorBrewer, rgdal, rasterVis, raster, sp, TTR

#To load the package "animation", you can do 
library(animation)

#You can also load all these packages in one shot
#using pacman
#install.packages("pacman")
library(pacman)
pacman::p_load(animation, chron, e1071, fields, ggplot2, lattice, 
            latticeExtra, maps, mapdata, mapproj, matrixStats, ncdf4, 
            NLRoot, RColorBrewer, rgdal, rasterVis, raster, sp, TTR)

#The zipped data file:
# https://www.cambridge.org/climatemathematics/data.zip

#On your computer, you can create a directory called 
#climmath under your user name. 
#The one used in the book is Users/sshen/climmath  
#You unzip the data and move the data folder under 
#the Users/sshen/climmath directory. 
#A data folder will created:  
#Users/sshen/climmath/data. 
#The data folder contains about 400 MB of data. 
#Place all the R codes in the directory  Users/sshen/climmath. 
#Then, you can run all the codes in this book after replacing sshen 
#by your user name on your own computer. 

```


