---
title: "&nbsp;"
output:
  html_document: default
  pdf_document: default
---

<center><h1>R Code for Climate Mathematics:</h1></center>
<center><h1>Theory and Applications</h1></center>
<p>&nbsp;</p>
<center><h3>A Cambridge University Press book By</h3></center>
<center><h3>SSP Shen and RCJ Somerville</h3></center>

<p>&nbsp;</p>
<p>&nbsp;</p>

<div align="center">Version 1.0 released in July 2019 and coded by Dr. Samuel Shen, Distinguished Professor</div>
<div align="center">San Diego State University, California, USA</div> 
<div align="center"><https://shen.sdsu.edu></div> 
<div align="center">Email: <sshen@sdsu.edu></div> 
<p>&nbsp;</p>
<div align="center">Version 2.0 compiled and updated by Momtaza Sayd</div> 
<div align="center">San Diego State University May 2021</div>
<div align="center">Version 3.0 compiled and updated by Joaquin Stawsky</div>
<div align="center">San Diego State University June 2022</div>


<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
## Chapter 7: Calculus Applications to Climate Science II: Integrals

### Plot Fig. 7.3: Corpus Christi: Elevation vs. F = g/(RT)
```{r}
setwd("~/sshen/climmath")
dat <- read.table("data/CorpsCristiTX13295R1.txt", header = FALSE)
dim(dat)
#plot.new()
da2 <- dat
for(i in 1:146){if(dat[i,4] >9999){da2[i,4] = NA} }
R <- 287.055 #[J/(kg K)]
g <- 9.8 #[m/sec^2]
  
par(mar = c(4.2,4.2,3,0.5))
plot(da2[,3],(g/R)/(273.15+da2[,4]/10),type = "l", 
     ylim = c(0,0.0002), lwd = 8,col = "red",
     xlab = "Elevation: z [m]", ylab = "F = g/(RT)",
     main = "Hypsometric Equation Integrand Based on the Radiosonde Data
     Corpus Christi, Texas, 0:00UTC, 12 September 2013",
     cex.axis = 1.2, cex.lab = 1.2)
lines(c(14, 9992),c((g/R)/(273.15+26.8),(g/R)/(273.15-34.1)), lwd = 2, col = "blue")
lines(c(14, 14),c(0,(g/R)/(273.15+26.8)), lwd = 2, col = "blue")
lines(c(9992, 9992),c(0,(g/R)/(273.15-34.1)), lwd = 2, col = "blue")
lines(c(14, 9992),c(0,0), lwd = 2, col = "blue")
text(1200,0.000075,"F1", cex = 1.2)
text(11500,0.000075,"F2", cex = 1.2)
text(1500,0.000007,"z1", cex = 1.2)
text(11500,0.000007,"z2", cex = 1.2)
points(c(14,10000,14,10000), 
       c(0,0,(g/R)/(273.15+26.8),(g/R)/(273.15-34.1)), 
       pch = 19, cex = 0.6)
```

<p>&nbsp;</p>
### Plot Fig. 7.5
```{r}
#Planck's law of blackbody radiation for the Sun and Earth
#setwd("/Users/sshen/climmath")
#Solar radiation
#setEPS()
#postscript("fig0705a.eps", height = 10, width = 8)
h <- 6.626070040
k <- 1.38064852
c <- 3
x <- seq(0.01,3,len = 2000)
et2 <- function(x){(2*h*c^2/x^5)/(exp(h*c/(k*x*T))-1)}
#plot(1, lwd = 0, xlim = c(0,3), ylim = c(0,30))
par(mar = c(4,5.0,1.5,0.5))
plot(1,type = "l",lwd = 0, xlim = c(0,3), ylim = c(0,30),
     main = "Planck's Law of Blackbody Radiation Around 5,000 K",
     xlab = expression(paste("Wavelength [", mu, "m]")),
     ylab = expression(paste("Spectral Power Flux of Radiation: ", "[kW/(", m^2, " nm)]")),
     cex.lab = 1.2, cex.axis = 1.2, cex.main = 1.3)
segments(0.41,0,0.41,30,lwd = 4,col = "purple")
segments(0.43,0,0.43,30,lwd = 6,col = "blue4")
segments(0.46,0,0.47,30,lwd = 10,col = "blue")
segments(0.52,0,0.52,30,lwd = 14,col = "green")
segments(0.57,0,0.57,30,lwd = 8,col = "yellow")
segments(0.61,0,0.61,30,lwd = 9,col = "orange")
segments(0.69,0,0.69,30,lwd = 24,col = "red")
text(1.1,-0.2, "infrared", cex = 1.2)
text(0.63,-0.2, "visible", cex = 1.2)
text(0.15,-0.2, "ultraviolet", cex = 1.2)
T <- 5.0
lines(x,et2(x), type = "l", col = "darkolivegreen2", lwd = 2)
T <- 5.772
lines(x,et2(x), type = "l", col = "blue4", lwd = 2)
T <- 4.0
lines(x,et2(x), type = "l", col = "brown3", lwd = 2)
legend(1.5,25, 
       legend = c("T = 5772 K","T = 5000 K", "T = 4000 K" ), 
       col = c("blue4","darkolivegreen2","brown3"),
       lty = 1, lwd = 3,
       cex = 1.2)
#dev.off()
```

<p>&nbsp;</p>
### Earth radiation
```{r}
#setEPS()
#postscript("fig0705b.eps", height = 10, width = 8)
par(mar = c(4,5.0,1.5,0.5))
h <- 6.626070040
k <- 1.38064852
c <- 3
et <- function(x){10^6*(2*h*c^2/x^5)/(exp(h*c/(k*x*T))-1)}
x <- seq(0.01,80,len = 2000)
T <- 0.288
plot(x,et(x), type = "l", col = "black", lwd = 2,
     main = expression(paste("Planck's Law of Blackbody Radiation Around 15",degree,"C")),
     xlab = expression(paste("Wavelength [", mu, "m]")),
     ylab = expression(paste("Spectral Power Flux of Radiation: ", 10^-6, "[kW/(", m^2, " nm)]")),
     cex.lab = 1.2, cex.axis = 1.2, cex.main = 1.3, ylim = c(0,10))
T <- 0.290
lines(x,et(x), type = "l", lty = 3, lwd = 2)
T <- 0.286
lines(x,et(x), type = "l", lty = 2, lwd = 2)
legend(40,9, 
       legend = c(expression(paste("T = 15+2 ",degree,"C")),
                  expression(paste("T = 15 ",degree,"C")),
                  expression(paste("T = 15-2 ",degree,"C"))), 
       #       col = c("red","black","blue"),
       lty = c(3,1,2), lwd = 3,
       cex = 1.2)
#dev.off()
```


