---
title: "&nbsp;"
output:
  html_document: default
  pdf_document: default
---

<center><h1>R Code for Climate Mathematics:</h1></center>
<center><h1>Theory and Applications</h1></center>
<p>&nbsp;</p>
<center><h3>A Cambridge University Press book By</h3></center>
<center><h3>SSP Shen and RCJ Somerville</h3></center>

<p>&nbsp;</p>
<p>&nbsp;</p>

<div align="center">Version 1.0 released in July 2019 and coded by Dr. Samuel Shen, Distinguished Professor</div>
<div align="center">San Diego State University, California, USA</div> 
<div align="center"><https://shen.sdsu.edu></div> 
<div align="center">Email: <sshen@sdsu.edu></div> 
<p>&nbsp;</p>
<div align="center">Version 2.0 compiled and updated by Momtaza Sayd</div> 
<div align="center">San Diego State University May 2021</div>
<div align="center">Version 3.0 compiled and updated by Joaquin Stawsky</div>
<div align="center">San Diego State University June 2022</div>

<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
## Chapter 4: Climate Data Matrices and Linear Algebra

### Matrix algebra
```{r}
#Matrix subtraction
matrix(c(1,1,1,-1), nrow = 2) - matrix(c(1,3,2,0), nrow = 2)

#Matrix multiplication
matrix(c(1,1,1,-1), nrow = 2) %*% matrix(c(1,2,3,4), nrow = 2)

matrix(c(1,2,3,4), nrow = 2) %*% matrix(c(1,1,1,-1), nrow = 2) 

#Find the inverse of a matrix
solve(matrix(c(1,1,1,-1), nrow = 2))

#Verify the inverse matrix 
matrix(c(0.5,0.5,0.5,-0.5), nrow = 2) %*% matrix(c(1,1,1,-1), nrow = 2) 


#Solve a group of linear equations
solve(matrix(c(1,1,1,-1),nrow = 2),c(20,4))
#This is the result x1 = 12, and x2 = 8.

#Compute a covariance matrix
A <- matrix(c(1,-1,2,0,3,1),nrow = 2)
A
covm <- (1/(dim(A)[2]))*A%*%t(A)
covm #is the covariance matrix.

#Matrix multiplication of a vector changes the vector direction 
u <- c(1,-1)
v <- covm%*%u
v
#u and v are in different directions.

#Matrix mutiplication of an eigenvector does change direction
ew <- eigen(covm)
eigen(covm)$values
eigen(covm)$vectors
#Verify the eigenvectors and eigenvalues
covm%*%ew$vectors[,1]/ew$values[1]

#This is the first eigenvector

#SVD: Decomposition and recovery of a space-time data matrix 
#Develop a 2-by-3 space-time data matrix for SVD
A <- matrix(c(1,-1,2,0,3,1),nrow = 2)
A
#Perform SVD calculation
msvd <- svd(A)
msvd
msvd$d
msvd$u
msvd$v
#One can verify that A = UDV', where V' is transpose of V.
verim <- msvd$u%*%diag(msvd$d)%*%t(msvd$v)
verim
round(verim)
#This is the original data matrix A

#Covariance matrix and its eigen-problem
covm <- (1/(dim(A)[2]))*A%*%t(A)
eigcov <- eigen(covm)
eigcov$values
eigcov$vectors

#Eigenvalues from SVD and from a covariance matrix
((msvd$d)^2)/(dim(A)[2])
eigcov$values
```

<p>&nbsp;</p>
### Plot Fig. 4.2
```{r}
#SOI and the Standardized SLP data at Darwin and Tahiti 
setwd("~/sshen/climmath")
#setEPS()
#postscript("fig0402a.eps", height = 4.9, width = 7)
#par(mar = c(4.0,4.2,1.5,0.5))
Pta <- read.table("data/PSTANDtahiti.txt", header = F)
# Remove the first column that is the year
ptamon <- Pta[,seq(2,13)]
#Convert the matrix into a vector according to mon: Jan 1951, Feb 1951, ..., Dec 2015
ptamonv <- c(t(ptamon))
#Generate time ticks from Jan 1951 to Dec 2015
xtime <- seq(1951, 2016-1/12, 1/12)
# Plot the Tahiti standardized SLP anomalies
plot(xtime, ptamonv,type = "l",xlab = "Year",
     ylab = "Pressure Anomalies", cex.lab = 1.2, cex.axis = 1.2,
     main = "Standardized Tahiti SLP Anomalies", col = "red",
     xlim = range(xtime), ylim = c(-4,4))
text(1952,3.5, "(a)", cex = 1.2)
# dev.off()

```

<p>&nbsp;</p>
### Do the same for Darwin SLP
```{r}
setwd("~/sshen/climmath/")
#setEPS()
#postscript("fig0402b.eps", height = 4.9, width = 7)
#par(mar = c(4.0,4.2,1.5,0.5))
Pda <- read.table("data/PSTANDdarwin.txt", header = F)
pdamon <- Pda[,seq(2,13)]
pdamonv <- c(t(pdamon))
plot(xtime, pdamonv,type = "l",cex.lab = 1.2, cex.axis = 1.2,
     xlab = "Year",ylab = "Pressure", 
     main = "Standardized Darwin SLP Anomalies", col = "blue",
     xlim = range(xtime), ylim = c(-4,4))
text(1952,3.5, "(b)", cex = 1.2)
# dev.off()
```

<p>&nbsp;</p>
### Plot the SOI index 
```{r}
#setEPS()
#postscript("fig0402c.eps", height = 4.9, width = 7)
#par(mar = c(4.0,4.2,2.0,0.5))
SOI <- ptamonv-pdamonv
plot(xtime, SOI ,type = "l", cex.lab = 1.2, cex.axis = 1.2,
     xlab = "Year",ylab = "SOI index", 
     col = "black",xlim = range(xtime), ylim = c(-6,6), lwd = 1)
#Add ticks on top edge of the plot box
axis (3, at = seq(1951,2015,4), labels = seq(1951,2015,4))
#Add ticks on the right edge of the plot box
axis (4, at = seq(-4,4,2), labels = seq(-4,4,2))
lines(xtime, rep(0,length(xtime)))
text(1985, 5, "SOI = Tahiti Pressure - Darwin Pressure", cex = 1.2)
text(1952,5.6, "(c)", cex = 1.2)
#abline(lm(SOI ~ xtime), col = "red", lwd = 2)
# dev.off()
```

<p>&nbsp;</p>
### CSOI and AMO time series comparison
```{r}
setwd("~/sshen/climmath/")
# setEPS()
#postscript("fig0402d.eps", height = 4.9, width = 7)
par(mar = c(4,4.2,1.6,4))
cnegsoi <- -cumsum(ptamonv-pdamonv)
#par(mgp = c(2,2,4,0))
plot(xtime, cnegsoi,type = "l", cex.lab = 1.2, cex.axis = 1.2,
     xlab = "Year",ylab = "Negative CSOI index", 
     main = "CSOI and AMO Index Comparison", 
     col = "purple",xlim = range(xtime), ylim = range(cnegsoi), lwd = 1.5)
legend(1960,15, col = c("purple"),lty = 1,lwd = 2.0,
       legend = c("CSOI"),bty = "n",text.font = 2,cex = 1.2)
text(1951,12, "(d)", cex = 1.2)
#AMO data and plot
amodat <- read.table("data/AMO1951-2015.txt", header = FALSE)
amots <- as.vector(t(amodat[,2:13]))
par(new = TRUE)
plot(xtime, amots,type = "l",col = "darkgreen",
     cex.lab = 1.2, cex.axis = 1.2,
     lwd = 1.5,axes = FALSE,xlab = "",ylab = "")
legend(1960,0.45, col = c("darkgreen"),lty = 1,lwd = 2.0,
       legend = c("AMO index"),bty = "n",text.font = 2,cex = 1.2)
#Suppress the axes and assign the y-axis to side 4
axis(4, cex.axis = 1.3)
mtext("AMO index",side = 4,line = 3, cex = 1.2)
# dev.off()
```

<p>&nbsp;</p>
### Compute the weighted SOI by SVD
```{r}
#Time-space data format
ptada <- cbind(ptamonv,pdamonv)
#Space-time data format
ptada <- t(ptada)
#Implement SVD for the sapce-time data
svdptd <- svd(ptada)
recontd <- svdptd$u%*%diag(svdptd$d[1:2])%*%t(svdptd$v)
recontd <- ptada
#Two EOFs from the space-time data
U <- svdptd$u
U
#These are the weights for the WSOI
svdptd$d
D <- diag(svdptd$d)
D

#PCs: The principal components
V <- svdptd$v
V
```

<p>&nbsp;</p>
### Plot Fig. 4.3
```{r}
U <- svdptd$u #The matrix of the EOF spatial patterns
V <- svdptd$v #The matrix of the PC temporal patterns
D <- diag(svdptd$d) #The diagonal matrix of standard deviations 
recontada <- U%*%D%*%t(V) #Verify the SVD decomposition
svd_error <- ptada - recontada
max(svd_error)
min(svd_error)
```

<p>&nbsp;</p>
### Plot WSOI1
```{r}
#setEPS()
#postscript("fig0403a.eps", height = 4.9, width = 7)
par(mar = c(4,4.2,2.0,0.5))
xtime <- seq(1951, 2016-1/12, 1/12)
wsoi1 <- D[1,1]*t(V)[1,]
plot(xtime, wsoi1,type = "l", cex.lab = 1.2, cex.axis = 1.2,
     xlab = "Year",ylab = "WSOI1", 
     col = "black",xlim = range(xtime), ylim = c(-4,4))
axis (3, at = seq(1951,2015,4), labels = seq(1951,2015,4))
text(1985,-3.7,"Weighted SOI1 Index", cex = 1.2)
text(1952,3.5, "(a)", cex = 1.2)
# dev.off()
```

<p>&nbsp;</p>
### Plot WSOI2
```{r}
#setEPS()
#postscript("fig0403b.eps", height = 4.9, width = 7)
par(mar = c(4,4.2,2.0,0.5))
xtime <- seq(1951, 2016-1/12, 1/12)
wsoi2 <- D[2,2]*t(V)[2,]
plot(xtime, wsoi2,type = "l", cex.lab = 1.3, cex.axis = 1.3,
     xlab = "Year",ylab = "WSOI2", 
     col = "blue",xlim = range(xtime), ylim = c(-4,4))
axis (3, at = seq(1951,2015,4), labels = seq(1951,2015,4))
text(1985,-3.7,"Weighted SOI2 Index", cex = 1.5)
text(1952,3.5, "(b)", cex = 1.3)
# dev.off()
```

<p>&nbsp;</p>
### Cumulative WSOIs
```{r}
setwd("~/sshen/climmath/")
#setEPS()
#postscript("fig0403c.eps", height = 4.9, width = 7)
par(mar = c(4,4.2,2.0,4.2))
cwsoi1 <- cumsum(wsoi1)
plot(xtime, cwsoi1,type = "l",cex.lab = 1.2, cex.axis = 1.2,
     xlab = "Year",ylab = "Weighted SOI1", 
     col = "black",lwd = 3, ylim = c(-200,50),
     main = "Comparison between CWSOI1 and the Smoothed AMO Index")
text(1951,40, "(c)", cex = 1.3)
#axis (3, at = seq(1951,2015,4), labels = seq(1951,2015,4))
legend(1970,20, col = c("black"),lty = 1,lwd = 3.0,
       legend = c("CWSOI1"),bty = "n",text.font = 2,cex = 1.2)
#Superimpose CSOI time series on this CWSOI1
cnegsoi <- -cumsum(ptamonv-pdamonv)
lines(xtime, cnegsoi,type = "l",col = "purple", lwd = 3.0)
legend(1970,2, col = c("purple"),lty = 1,lwd = 3.0,
       legend = c("CSOI"),bty = "n",text.font = 2,cex = 1.2)
#24-month ahead moving average of the monthly AMO index
amodat <- read.table("data/AMO1951-2015.txt", header = FALSE)
amots <- as.vector(t(amodat[,2:13]))
#install.packages("TTR")
library("TTR")
amomv <- SMA(amots,n = 24, fill = NA) 
#Average of the previous n points
par(new = TRUE)
xtime <- seq(1951,2015,len = 780)
plot(xtime, amomv[37:816],type = "l",col = "darkgreen",
     lwd = 3,axes = FALSE,xlab = "",ylab = "",
     cex.lab = 1.2, cex.axis = 1.2,
     ylim = c(-0.5, 0.5))
legend(1970,0.23, col = c("darkgreen"),lty = 1,lwd = 3.0,
       legend = c("AMO index"),bty = "n",text.font = 2,cex = 1.2)
#Suppress the axes and assign the y-axis to side 4
axis(4, cex.axis = 1.2, col.axis = "darkgreen")
mtext("Smoothed AMO index",
      side = 4,line = 3, cex = 1.2, col = "darkgreen")
# dev.off()

```

<p>&nbsp;</p>
### Plot cumulative WSOI2: CWSOI2
```{r}
#setEPS()
#postscript("fig0403d.eps", height = 4.9, width = 7)
par(mar = c(4,4.2,2.0,0.5))
cwsoi1 <- cumsum(wsoi1)
wsoi2 <- D[2,2]*t(V)[2,]
cwsoi2 <- cumsum(wsoi2)
plot.new()
# par(mar = c(4,4,3,4))
plot(xtime, cwsoi2,type = "l",xlab = "Year",ylab = "CWSOI2", 
     cex.lab = 1.2, cex.axis = 1.2,
     col = "blue",lwd = 3, ylim = c(-200,50),
     main = "CWSOI2 Index: The Cumulative PC2")
text(1951,40, "(d)", cex = 1.3)
# dev.off()
```

<p>&nbsp;</p>
### Plot Fig. 4.4
```{r}
#Display the two ENSO modes on a world map
library(maps)
library(mapdata)

plot.new()
par(mfrow = c(2,1))

par(mar = c(0,0,0,0)) #Zero space between (a) and (b)
map(database = "world2Hires",ylim = c(-70,70), mar = c(0,0,0,0))
grid(nx = 12,ny = 6)
points(231, -18,pch = 16,cex = 2, col = "red")
text(231, -30, "Tahiti 0.61", col = "red")
points(131, -12,pch = 16,cex = 2.6, col = "blue")
text(131, -24, "Darwin -0.79", col = "blue")
axis(2, at = seq(-70,70,20),
     col.axis = "black", tck = -0.05, las = 2, line = -0.9,lwd = 0)
axis(1, at = seq(0,360,60),
     col.axis = "black",tck = -0.05, las = 1, line = -0.9,lwd = 0)
text(180,30, "El Nino Southern Oscillation Mode 1",col = "purple",cex = 1.3)
text(10,-60,"(a)", cex = 1.4)
box()

# par(mar = c(0,0,0,0)) #Plot mode 2
map(database = "world2Hires", ylim = c(-70,70), mar = c(0,0,0,0))
grid(nx = 12,ny = 6)
points(231, -18,pch = 16,cex = 2.6, col = "red")
text(231, -30, "Tahiti 0.79", col = "red")
points(131, -12,pch = 16,cex = 2, col = "red")
text(131, -24, "Darwin 0.61", col = "red")
text(180,30, "El Nino Southern Oscillation Mode 2",col = "purple",cex = 1.3)
axis(2, at = seq(-70,70,20),
     col.axis = "black", tck = -0.05, las = 2, line = -0.9,lwd = 0)
axis(1, at = seq(0,360,60),
     col.axis = "black",tck = -0.05, las = 1, line = -0.9,lwd = 0)
text(10,-60,"(b)", cex = 1.4)
box()
```

<p>&nbsp;</p>
### An example of multivariate regression
```{r}
x1 <- c(1,2,3) #Given the coordinates of the 3 points
x2 <- c(2,1,3)
y <- c(-1,2,1)
df <- data.frame(x1,x2,y) #Put data into the data.frame format
fit <- lm(y ~ x1 + x2, data = df)
fit #Show the regression results

1.667*x1-1.333*x2 #Verify that 3 points determining a plane
```

<p>&nbsp;</p>
### Multivariate regression with four data points
```{r}
u <- c(1,2,3,1)
v <- c(2,4,3,-1)
w <- c(1,-2,3,4)
mydata <- data.frame(u,v,w)
myfit <- lm(w ~ u + v, data = mydata)
summary(myfit) #Show the result
```

<p>&nbsp;</p>
### Another example of multivariate regregression
```{r}
dat <- matrix(rnorm(40),nrow = 10, dimnames = list(c(letters[1:10]), c(LETTERS[23:26])))
fdat <- data.frame(dat)
fit <- lm(Z ~ W + X + Y, data = fdat)
summary(fit)
```
