---
title: "&nbsp;"
output:
  html_document: default
  pdf_document: default
---

<center><h1>R Code for Climate Mathematics:</h1></center>
<center><h1>Theory and Applications</h1></center>
<p>&nbsp;</p>
<center><h3>A Cambridge University Press book By</h3></center>
<center><h3>SSP Shen and RCJ Somerville</h3></center>

<p>&nbsp;</p>
<p>&nbsp;</p>

<div align="center">Version 1.0 released in July 2019 and coded by Dr. Samuel Shen, Distinguished Professor</div>
<div align="center">San Diego State University, California, USA</div> 
<div align="center"><https://shen.sdsu.edu></div> 
<div align="center">Email: <sshen@sdsu.edu></div> 

<p>&nbsp;</p>
<div align="center">Version 2.0 compiled and updated by Momtaza Sayd</div> 
<div align="center">San Diego State University May 2021</div>
<div align="center">Version 3.0 compiled and updated by Joaquin Stawsky</div>
<div align="center">San Diego State University June 2022</div>

<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
## Chapter 6: Calculus Applications to Climate Science I: Derivatives

### Plot Fig. 6.1
```{r}
#setEPS()
#postscript("fig0601.eps", height = 8, width = 10)
#Cess' Budyko parameters
par(mar = c(4.5,5,2,0.5))
A <- 212
B <- 1.6
ep <- 0.6
sg <- 5.670373*10^(-8)
T <- seq(-50,50, by=0.1)
S <- 1365
alf <- 0.30
plot(T,A+B*T, xlim = c(-50,50), ylim = c(0,500), 
     type = "l",lwd = 4, cex.lab = 1.2, cex.axis = 1.2,cex.main = 1.35,
     main = "Radiation Law and Its Approximations",
     xlab = expression(paste("Temperature T [",degree,"C]")),
     ylab = expression(paste("Outgoing Radiation: ", E[out], " [W/",m^2,"]")))
lines(T, ep*sg*(273.15+T)^4,col = "blue", lwd = 3)
lines(T, 189.4 + 2.77*T,type = "l", col = "purple",lwd = 3)
lines(T, (1-alf)*S/4*rep(1, length(T)), col = "red",lwd = 2)
legend(-50, 510, legend = c("Budyko radiation formula", 
                          "Stefan-Boltzmann radiation law",
                          "Linear Approximation of SB law",
                          "Observed outgoing radiation"),
       col = c("black", "blue","purple","red"),
       lty = 1, bty = "n",lwd = 3, cex = 1.2)
text(-40,180, "Budyko formula",cex = 1.2)
text(15,310, "Stefan-Boltzmann law", col = "blue",cex = 1.2)
text(-3,65, "Linear Approximation to Stefan-Boltzmann law", 
     col = "purple",cex = 1.2)
text(-37,255, expression(paste( E[out], " = 239 [W/",m^2,"]")), 
     col = "red", cex = 1.2)
#dev.off()
```

<p>&nbsp;</p>
### Generate Table 6.1
```{r}
x <- seq(-0.3,0.3, by = 0.1)
fx <- c(1:7)
lx <- c(1:7)
ex <- c(1:7)
for(i in 1:7){fx[i] <- (1+x[i])^4
  lx[i] <- 1+4*x[i]
  ex[i] <- ((1+x[i])^4-(1+4*x[i]))/((1+x[i])^4)*100
}
round(cbind(x, fx, lx, ex), digits = 4)
```

<p>&nbsp;</p>
### Solve an equation by the bisection method
```{r}
#install.packages("NLRoot") #if not installed before
library(NLRoot)
func <- function(x){x^3-x-1}
#[1] 1.324716 is the solution
BFfzero(func, 0, 2)
```

<p>&nbsp;</p>
### Another bisection example
```{r}
f1 <- function(x){(1+x)^4-(2+x)}
#[1] 0.2207428 is the solution
BFfzero(f1,0,2)
```

<p>&nbsp;</p>
### Solve an EBM by the bisection method
```{r}
S <- 1368
ep <- 0.6
sg <- 5.670373*10^(-8)
f2 <- function(T){ep*sg*T^4 - (1-(0.5 - 0.2*tanh((T-265)/10)))*(S/4)}
#[1] 289.3097 is the solution
BFfzero(f2,270,300)
```

<p>&nbsp;</p>
### R code for the Newton's method to solve an equation
```{r}
newton <- function(f, tol = 1E-12,x0 = 1,N = 20) {
  h <- 0.001
  i <- 1; x1 <- x0
  p <- numeric(N)
  while (i <= N) {
    df.dx <- (f(x0+h)-f(x0))/h
    x1 <- (x0 - (f(x0)/df.dx))
    p[i] <- x1
    i <- i + 1
    if (abs(x1-x0) < tol) break
    x0 <- x1
  }
  return(p[1:(i-1)])
}
f <- function(x) { x^3 + 4*x^2 -10 }
root <- newton(f, tol = 1E-12, x0 = 1, N = 10)
root
uniroot(f,c(1,2), tol = 10^(-6))

```

<p>&nbsp;</p>
### Plot Fig. 6.2
```{r,fig.width = 8, fig.height = 6}
#Illustration of Newton's method
#setEPS()
#postscript("fig0602.eps", height = 6, width = 8)
par(mar = c(0,1,0,0))
x <- seq(0.2,1.7, len = 30)
f <- function(x) { x^3 + 4*x^2 -10 }
g <- function(x) { 3*x^2 + 8*x }
plot(x,f(x), type = 'l', lwd = 1.5, bty = "n", 
     xaxt = "n",yaxt  = "n", xlab = "", ylab = "",
     xlim = c(0,1.8),ylim = c(-10,7))
axis(1, at = c(0, 0.5, 1.0, 1.5, 1.7), pos = 0, cex.lab = 1.3)
axis(2, at = c(-8, -6, -4, -2, 0, 2, 4), pos = 0, cex.lab = 1.3, las = 1)
arrows(0,0, 1.8,0, angle = 10, length = 0.25)
arrows(0,-11, 0,7, angle = 10, length = 0.25)
text(1.8, 0.5, expression(x), cex = 1.4)
text(0.05, 7, expression(y), cex = 1.4)
text(0.4, -8, expression(y == x^3 + 4*x^2 -10), cex = 1.4) 
#Plot the initial points and then follow tangent lines
x0 <- 1.0
x1 <- 1.4543
x2 <- 1.3689
lines(x,f(x0) + g(x0)*(x - x0),type = "l", lty = 2, col = "red")
text(0.73,-8.5, expression(paste("Tangent line through ", P[0])),
     col = "red", cex = 1.2, srt = 42)
points(1,0, pch = 19)
text(1.0, 1.3, "Initial guess", cex = 1.4, col = "blue")
text(1.0, 0.5, expression(x[0]), cex = 1.4, col = "blue")
arrows(1,0,1,-5, angle = 10, length = 0.15, col = "blue")
points(x0, f(x0), col = "red", pch = 20)
text(x0+ 0.02, f(x0) - 0.5, expression(P[0]), cex = 1.4, col = "red")
arrows(x0,f(x0), x1,0, angle = 10, length = 0.15, col = "blue")
points(x1,0, pch = 19)
text(x1+0.04, 0.3, expression(x[1]), cex = 1.4, col = "blue")
arrows(x1,0,x1,f(x1), angle = 10, length = 0.15, col = "blue")
points(x1, f(x1), col = "red", pch = 20)
text(x1+ 0.05, f(x1), expression(P[1]), cex = 1.4, col = "red")
lines(x,f(x1) + g(x1)*(x - x1),type = "l", lty = 2, col = "red")
text(1.05,-6.5, expression(paste("Tangent line through ", P[1])),
     col = "red", cex = 1.2, srt = 56)
arrows(x1,f(x1),x2,0, angle = 10, length = 0.15, col = "blue")
points(x2,0, pch = 19)
text(1.34,0.5, expression(x[2]), cex = 1.4, col = "blue")
text(0.8, 6, "Newton's method for finding a root:", cex = 1.8)
text(0.8, 5,"Follow the blue arrows from the initial guess", 
     cex = 1.3, col = "blue")
text(0.8, 4, expression(x[0] %=>% P[0] %=>% x[1] %=>% P[1] %=>% x[2] %=>% ...), 
     cex = 1.3, col = "blue")                            
#dev.off()
```

<p>&nbsp;</p>
### Use Newton's method to solve an equation
```{r}
f <- function(x) { (x+1)^4 -(2+x) }
root <- newton(f, tol = 1E-12, x0 = 1, N = 10)
root
```

<p>&nbsp;</p>
### Find multiple solutions of an EBM
```{r}
S <- 1365
ep <- 0.6
sg <- 5.670373*10^(-8)
#Define the function for energy
f3 <- function(T){return(ep*sg*T^4 - 
                           (1-(0.5 - 0.2 * tanh ((T-265)/10)))*(S/4))}
root1 <- newton(f3, tol = 1E-12, x0 = 220,N = 20)
root1
root2 <- newton(f3, tol = 1E-12, x0 = 270,N = 20)
root2
root3 <- newton(f3, tol = 1E-12, x0 = 300,N = 20)
root3
root5 <- newton(f3, tol = 1E-12, x0 = 100, N = 20)
root5
```

<p>&nbsp;</p>
### Use R to find derivatives of a function symbolically
```{r}
D(expression(x^2),"x")
# 2 * x

D(expression(exp(-x^2)),"x")
# -(exp(-x^2) * (2 * x))

D(expression(sin(-3*t)-2*cos(4*t-0.3*pi)),"t")
# -(cos(-3 * t) * 3 - 2 * (sin(4 * t - 0.3 * pi) * 4))

D(expression(-g*t^2/2 + v0*t + h0),"t")
# v0 - g * (2 * t)/2
# Find derivative of this result function to find the second derivative
D(expression(v0 - g * (2 * t)/2),"t")
# -(g * 2/2)
#or simply
D(D(expression(-g*t^2/2 + v0*t + h0),"t"),"t")
# -(g * 2/2)
#The third-order derivative
D(D(D(expression(-g*t^2/2 + v0*t + h0),"t"),"t"),"t")
```
